Ensemble Stability of Biological Networks

### 2014-09-22 paper.tex
update title
edit introduction

% Title must be 150 characters or less
\begin{center}
{\Large
\textbf{Ensemble Stability of Biological Networks}
}
% Insert Author names, affiliations and corresponding author email.
\\[.5cm]
Aviv Bergman$^{1,2,3,4}$,
Raymond S. Puzio$^{1}$,
Cameron Smith$^{1}$,
\\[.5cm]
$^1$Department of Systems and Computational Biology,\\
$^2$Dominick P. Purpura Department of Neuroscience,\\
$^3$Department of Pathology, Albert Einstein College of Medicine,\\
1301 Morris Park Ave, Bronx, NY 10461, USA\\
$^4$Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA
\\[.5cm]
% $\ast$To whom correspondence should be addressed; E-mail: aviv@einstein.yu.edu.
\end{center}

@@

The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to consider a particular example, derive a system of differential equations to model that example, try to fit the model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved. Recent work has demonstrated that as a result of sloppiness in the dependence of qualitative dynamic phenomena on the geometry of parameter space that this approach allows for a large variety of models to ``fit the data'' \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014}. In the face of uncertainty about the structure of such biological networks, to model the components under consideration as randomly interlinked becomes a reasonable approximation. This approach enables one to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}.

@@

Here we build on work aimed at investigating the stability of randomly connected dynamical systems to structural perturbations. We precisely compute the probability of stability as a distribution over system connectivity for all systems containing two or three interacting components. We then proceed to determine the probability of \emph{structural stability} or \emph{robustness} which is the distribution of stability to perturbation in the system structure for systems which are determined to be stable \cite{Smale1967}. This quantity is of interest at all scales of the biological hierarchy, and has been previously addressed in the study of the robustness of gene-regulatory networks \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}. We find that stability to structural perturbations of this class of dynamical systems is correlated with connectivity, number of cycles, and the number of links between strongly connected components of the underlying interdependency graph. The latter derives from the fact that the permutation of strongly connected components is a symmetry of system robustness. This work clarifies a thread of research that is extremely important for the continuing development of systems biology.

>>>>

% Title must be 150 characters or less
\begin{center}
{\Large
\textbf{Hierarchical Network Structure Promotes Dynamical Robustness}
}
% Insert Author names, affiliations and corresponding author email.
\\[.5cm]
Aviv Bergman$^{1,2,3,4}$,
Raymond S. Puzio$^{1}$,
Cameron Smith$^{1}$,
\\[.5cm]
$^1$Department of Systems and Computational Biology,\\
$^2$Dominick P. Purpura Department of Neuroscience,\\
$^3$Department of Pathology, Albert Einstein College of Medicine,\\
1301 Morris Park Ave, Bronx, NY 10461, USA\\
$^4$Santa Fe Institute, 1399 Hyde Park Road, Santa Fe, NM 87501, USA
\\[.5cm]
% $\ast$To whom correspondence should be addressed; E-mail: aviv@einstein.yu.edu.
\end{center}

@@

The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to consider a particular example, derive a system of differential equations to model that example, try to fit the model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. Recent work has demonstrated that as a result of sloppiness in the dependence of qualitative dynamic phenomena on the geometry of parameter space that this approach allows for a large variety of models to fit the data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In the face of uncertainty about the structure of such biological networks, to model the components under consideration as randomly interlinked becomes a reasonable approximation. This approach enables one to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}.

@@

Here we build on work aimed at investigating the stability of randomly connected dynamical systems to structural perturbations. We precisely compute the probability of stability as a distribution over system connectivity for all systems containing two or three interacting components. We then proceed to determine the probability of \emph{robustness} or \emph{structural stability} which is the distribution of stability to perturbation in the system structure for systems which are determined to be stable \cite{Smale1967}. This quantity is of interest at all scales of the biological hierarchy, and has been previously addressed in the study of the robustness of gene-regulatory networks \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}. We find that stability to structural perturbations of this class of dynamical systems is correlated with connectivity, number of cycles, and the number of links between strongly connected components of the underlying interdependency graph. The latter derives from the fact that the permutation of strongly connected components is a symmetry of system robustness. This work clarifies a thread of research that is extremely important for the continuing development of systems biology.
